---
title: "Packages tigris and ACS and plotting school district data."
output: html_document
---
R provides easy to use packages that wrap the census TIGER shapefile database and the census American Community Survey database. Because of this, it is easy to plot reasonable maps in R of descriptive statistics at the district level for a given state.

Furthermore, with a little bit of object hacking, you can create a geojson file useable in Tableau for your own purposes.
```{r needed_libs, message=FALSE}
library(rgdal) # Used for outputting geojson files, and a dependency for tigris
library(tigris) # Intertface for the TIGER database
library(acs) # Interface for the ACS API
library(rgeos) # For cropping
library(raster) # For cropping
library(rgdal) # For outputting geoJSON -- Not actually executed in RMD

```
You will need to request an API key from the census bureau for ACS API integration at:

  http://api.census.gov/data/key_signup.html

# Downloading School District Shapefiles

To download school district shapefiles we use the
```{r, eval = FALSE}
school_districts()
```
function from tigris.

Two arguments are required: State, which takes 2-letter state abbreviations, and type

Three kinds of school district shapefiles exist in TIGER:

* Unified districts (type = "unified")
* Elementary districts (type = "elementary")
* Secondary districts (type = "secondary")

To get a basic unified district shapefile for Mass.:
```{r basic_district, messages = FALSE}
MA_unified <- school_districts(state = "MA",
                               type = "unified")
```

# Hacking school district shapefiles together.
If you plot the unified districts, you won't get what you expect, exactly:

```{r unified_plot}
plot(MA_unified)
```

We will need to bring in the secondary and elementary districts to get the expected plot.
```{r sec_elem, messages = FALSE}
MA_secondary <- school_districts(state = "MA",
                                 type = "secondary")

MA_elementary <- school_districts(state = "MA",
                                  type = "elementary")
plot(MA_secondary)
plot(MA_elementary)
```

Sadly, for the time being, the built-in tigris function
```{r, eval = FALSE}
rbind_tigris()
```
to handle shapefile merging doesn't work with multiple tigris shapefile types, which different districts technically are (there is a pull request in to change this.)
```{r no_rbind_tigris, error = TRUE}
test <- rbind_tigris(MA_elementary, MA_secondary)
```
For the time being, we can hack around this by:

  1. Changing the tigris attribute for each shape file object.
  2. Renaming the "SDLEA"" columns for tigris_dataframe@data.
```{r hack_tigris}
attr(MA_unified, "tigris") <- "sdall"
attr(MA_secondary, "tigris") <- "sdall"
attr(MA_elementary, "tigris") <- "sdall"

names(MA_unified@data)[2] <- "SDLEA"
names(MA_secondary@data)[2] <- "SDLEA"
names(MA_elementary@data)[2] <- "SDLEA"

MA_all_sd <- rbind_tigris(MA_unified,
                          MA_secondary,
                          MA_elementary)

plot(MA_all_sd)
```
Still not quite there -- There's those polygons over water.
Peeking inside the @data gives a clue and we have a way to remove those polygons.
```{r remove water}
names(MA_all_sd@data)
                
# Any with no LAND area?
MA_all_sd@data[which(MA_all_sd@data$ALAND == "0"), ]

# Remove them.
MA_all_sd <- MA_all_sd[which(MA_all_sd@data$ALAND != "0"), ]

plot(MA_all_sd)
```
Still not quite there.

A county shapefile from tigris using the catrographic boundaries (cb=TRUE) to crop down to the land will result in what we want.
Cartographic boundaries isn't callable on school districts, but for areas that it is, it gives outlines without land.
```{r, messages = FALSE, warning = FALSE}
outline <- counties(state = "MA",
                    cb = TRUE, 
                    resolution = "500k", 
                    detailed = TRUE)

outline <- gUnaryUnion(outline)

dim(MA_all_sd)

MA_all_sd_land <- intersect(MA_all_sd, outline)

dim(MA_all_sd_land)

plot(MA_all_sd_land)
```

@Data is maintained:

```{r}
head(MA_all_sd_land, n = 2)
```

Final step before we dump this badboy into Tableau is outputting to geoJSON
```{r}
# writeOGR(MA_all_sd_land, "MA_all_sd_land.geojson", layer="MA_all_sd_land", driver="GeoJSON")
```

Check out
[the geoJSON dumped to Tableau Public](https://public.tableau.com/views/Pretty-fiedMASDs_/Sheet1?:embed=y&:display_count=yes)
for the result!

Sources

[tigris by Dr. Kyle Walker via CRAN](https://cran.r-project.org)

[Stack Overflow user cengel for the cropping](https://gis.stackexchange.com/a/173263)

[For ggplot2 mapping and easier ACS plotting](http://zevross.com/blog/2015/10/14/manipulating-and-mapping-us-census-data-in-r-using-the-acs-tigris-and-leaflet-packages-3/)